Edge sign prediction in general graphs and Correlation Clustering

### 2017-09-29 intro/draft.tex
first finished version of the edspi report

Graph theory has been around for almost three centuries~\autocite{biggs1976graph}, and it provides a
conceptually simple yet immensely rich framework to model situations where entities are connected
with each other~\autocite{ComplexNetworksApp11}. Coupled with the increasing availability of large
amount of relational data, it has recently spurred a lot of interest across various and tangible
applications:
\begin{itemize}[nosep,leftmargin=*]
  \item community detection ~\autocite{FortunatoSurvey10} Describe the task in one sentence and give
    some nice examples from the survey, if possible supported by real-world numbers;
  \item semi-supervised learning ~\autocites{SSL06}{graphSSL14} Idem
  \item node classification ~\autocite{nodeClassif11} Idem
  \item node embedding ~\autocite{representationLearning17} Idem
  \item link prediction ~\autocite{linkPredSurvey16} Idem
  \item influence maximization ~\autocites{infmaxKempe15}{infmax13} Idem
  \item network evolution ~\autocite{networkEvolution14} Idem
\end{itemize}

@@

Finally, we consider the case where there are more than two possible type of relations, which is
often called \emph{multilayer graphs}~\autocites{Kivela2014}{multiSurvey14}. In addition, we assume
that the graph is attributed, that is every node is described by a vector of numeric features called
the \emph{profile} of the node. Our assumption is then that links cannot be fully described by a
global homophily. Thus we take a more nuanced view, combining partial homophily (that is, nodes are
connected when they are similar on a subset of the attributes) and heterophily (that is, nodes are
connected when they are dissimilar on a subset of the attributes). As examples of the latter, think
of dating websites ---where most users are linked with users of the opposite
gender~\autocites{homophilyMyspace09}{Tinder16}; spreading innovation ---where meeting people with
different background and point of view is crucial to favor diversity and
creativity~\autocite{rogers2003diffusion}; and online news consumption ---where connecting people
from different sides of the political spectrum helps avoiding echo chambers and fueling a democratic
debate~\autocite{balancedNews17}. More generally, multilayer graphs have applications in various
fields. In social networks, where players of the Pardus massively multilayer online game can have
six types of relations (friendship, enmity, private message, trading, aggression and
bounty)~\autocite{Szell2010} or where photo sharers of the Flickr website can interact in eleven
ways (either directly or through comments, tags, groups and so on)~\autocite{RecoFlickrMulti11}. In
citation networks, where authors of the DBLP dataset are connected if they have co-authored a paper
in one the \np{1000} conferences that form the edge type~\autocite{communityDBLPbyConf05}, or where
\np{5000} SIAM papers can be connected for five reasons (abstract similarity, title similarity,
keywords similarity, author similarity or citation)~\autocite{articlesMultiSim11} in economic
networks, where 951 ports are connected by 3 kind of ships (bulk dry carriers, container ships and
oil tankers)~\autocite{ports3kindofships10} or in the International Trade Network, where 162
countries are connected by 96 kind of commodities they can exchange~\autocite{worldTradeNetwork10}.
And in biology, where genes co-expression network under 130 different experimental
conditions~\autocite{bioLayerExp11}

Existing works ~\autocite{egoNips12}

The social circles model defines a generative for connection inside ego
networks~\autocites{LeskovecEgo12}{LeskovecEgo14}. Basically, each node makes friends along $K$
categories and we learn the parameters that maximize the probability for each edge to belong to one
of the $K$ categories\footnote{But those categories can change from node to node.}.

Yixiang Fang, Reynold Cheng, Siqiang Luo, and Jiafeng Hu. 2016. Effective community search for large
attributed graphs. Proc. VLDB Endow. 9, 12 (August 2016), 1233-1244. DOI:
\url{http://dx.doi.org/10.14778/2994509.2994538}

Look for attributed community,  which are subgraphs satisfying both structure cohesiveness (i.e.,
its vertices are tightly connected) and keyword cohesiveness (i.e., its vertices share common
keywords),  enabling a better understanding of how and why a community is formed
To enable efficient AC search, we develop the CL-tree index structure and three algorithms based on
it. We evaluate our solutions on four large graphs, namely Flickr, DBLP, Tencent, and DBpedia. Our
results show that ACs are more effective and efficient than existing community retrieval approaches.

Find partitions by maximizing the densite of intra community edges, but weight the edge between $i$
and $j$ by a function that depends both of the $i$ and $j$ profile similarities and a learned weight
vector per community~\autocite{ZhangModelFree16}. Solved by alternating optimizing over node
assignment and over community weight learning.
vector

each node can be seen as a point in $R^d$ and because the dimensions are assumed to be not
independent, each point belong to a subspace. It's a well studied problem~\autocite{SCSurvey11}

Recovering $d$ different metric spaces whose intersection form the observed social multiplex in
almost linear time with bounded distortion~\autocite{Abraham2012a}. Assume there are $K$ social
categories modelled by Euclidean spaces $\mathcal{D}_i$. Users in there are near uniformly
distributed and categories have small local correlation:
\[ \forall i\neq i', \; \forall r, r' = O(polylog(n)),\,
  \forall u, u',\, |\mathcal{B}_i(u, r) \cap \mathcal{B}_{i'}(u', r')| \leq O(\log n) \]
These spaces give rise to small world networks $\mathcal{G}_i$ with edge probability $\propto
\mathcal{D}_i(u, v)^{-d}$ and we observe the real network $\mathcal{G} =\bigcup_i \mathcal{G}_i$.
From $\mathcal{G}$, the proposed algorithm recovers in $n \mathrm{polylog} n$ time a bounded
approximation $\mathcal{D}'_i$ of all $\mathcal{D}_i$
\[ \sigma \mathcal{D}_i(u, v) \leq \mathcal{D}'_i(u, v) \leq \delta \mathcal{D}_i(u, v) + \Delta \]

As hinted by the title, \autocite{Ahmed2016}\footnote{published version in
\autocite{ahmed2017roles}?} is a specific instantiation of a more general framework
about Role Discovery in Networks by the same authors. Originally, two nodes were said to have the
same role if they were structurally equivalent (that is, connected to the rest of the network in
identical ways). Hereafter, structural equivalence is relaxed into mere similarity. Furthermore, the
entities under consideration are now edges instead of nodes.
The framework has two main steps:
1) Automatically extracting a feature matrix X from an attributed graph
2) Applying clustering or factorization methods on X to assign roles to edges

The first step is mostly based on counting the number of k-motifs in which edges is involved,
although the authors stresses that their method could easily be extended to handle features derived
from node attributes. Letting m be the number of edges, the output of this step is a $m\times f$
matrix X of edge features.

In the second step, the authors seek a low rank approximation of X. Namely, they
want to minimize the Bregman divergence between X and UVT, where U and V are
two non-negative matrix of rank at most r. The optimization is carried out by a
parallel coordinate descent method called PCMF-BASIC and described in a separate
journal article: Parallel collective factorization for modeling large heterogeneous
networks.
There is also a regularization term, which is not given explicitly (although it
might have something to do with ensuring UVT sparsity). The rank r is chosen
dynamically through a Minimum Description Length approach.

The article ends with an extension to time evolving graph and some experimental
illustrations on the Enron mail dataset (and a citation one I assume). They also
give an example of application: detecting malicious user in Social Network. While its
node and most of its edges are normal, some edges might not be (harassment,
scam, …)

Stattner, Erick, and Martine Collard. "Link Clustering for Extracting Collaborative Patterns in a Scientific Co-Authored Network.", FAB workshop, ANOSAM 2017
they use their method of conceptual links
Groups of nodes are first created when they have common attributes. The set of links between two
groups is called a conceptual link in the sense that it corresponds to a relationship between two
sets of attributes, namely two concepts in the field of formal concept analysis [7]. The number of
links between two groups is then evaluated and when the frequency is greater than a given support
threshold $\beta$, we talk about frequent conceptual link. Note that we only keep maximal frequent conceptual links (mfcl), i.e. those who are not included in others.
Finally, the set of maximal frequent conceptual links is used to create a new network structure called a conceptual view which summarizes all the knowledge extracted from the initial network. In a conceptual view, each node corresponds to a set of attributes (in this context we call them Meta-nodes) and a link corresponds to maximal frequent conceptual links.
A formal presentation of frequent conceptual links can be found in [18].

Our approach is to seek a small number of vectors that, once assigned to every edge, best explain
the graph (\ie maximize a score function between the edge vector and the profiles of its endpoints).
From this initial formulation, we derive two optimization problems, and initially solve them on
synthetic data, generated with a few natural topological constraints. We compared these solutions
with the result of a baseline approach that simply clusters the pairwise profile vectors using
$k$-means.

>>>>

Graph theory has been around for almost three centuries~\autocite{biggs1976graph}, and it provides a
conceptually simple yet immensely rich framework to model situations where entities are connected
with each other~\autocite{ComplexNetworksApp11}. Coupled with the increasing availability of large
amount of relational data, it has recently spurred a lot of interest across various and tangible
applications:
\begin{itemize}[nosep,leftmargin=*]
  \item community detection~\autocite{FortunatoSurvey10}, where the goal is to cluster nodes in
    tightly connected group that are loosely connected with the rest of the graph, in order to
    better understand their relationships and the graph organization. Is has been used for instance
    to identify proteins functional groups~\autocite{clusterBio03} or see how different scientific
    fields relate based on publication data~\autocite{clusterScience08}.
  \item semi-supervised learning~\autocites{SSL06}{graphSSL14}, where during training, we are
    provided with unlabeled data at the same time as labeled data. Connecting similar instances
    allow propagating information along the graph, with applications in classifying text
    documents~\autocite{sslText09} or aligning categories and relations across multiple knowledge
    base~\autocite{sslKB13}.
  % \item node classification~\autocite{nodeClassif11} Idem
    % could add visualisation (Roberto Tamassia. 2013. Handbook of Graph Drawing and Visualization.
    % CRC Press.) but it's not exactly learning
  \item node embedding~\autocite{representationLearning17}, where we seek a low dimensional
    representation of nodes based on their structural patterns, usually in an unsupervised way,
    although it is also possible to include problem supervision when available. Such representation can
    then be used in downstream tasks, for instance visualisation~\autocite{LINE15} or the
    aforementioned node classification, even when new nodes join the graph after
    training~\autocite{inductiveRepresentation17}.
  \item link prediction~\autocite{linkPredSurvey16}, where the goal is, given a snapshot of the
    graph at time $t$, to return a set of non-existing links that will be created by time
    $t+\Delta$. Most methods are based on the assumption that link creation is driven by node
    similarity and it has been successfully applied to inferring potential interaction between
    proteins without expensive experiments~\autocite{linkPredBio06} and uncovering hidden
    association in criminal networks~\autocite{linkPredCrime08}.
  \item information and influence propagation~\autocite{infmax13}, which is the study of processes
    by which content is spread across networks, and how such processes can be influenced to speed
    them up or slow them down. Two main applications are selecting the best seed in a social
    network to promote a viral marketing campaign~\autocite{infmaxKempe15} and better contain the
    diffusion of actual biological viruses~\autocite{influenceBio13}.
  \item network evolution~\autocite{networkEvolution14}, which focuses on the mechanisms and
    consequences of the growth of most networks, as well as how to keep the results on some data
    mining algorithms \enquote{fresh} and relevant. Monitoring the changes in the interactions of
    proteins can indeed be used as an early indicator for some kind of
    diseases~\autocite{evolBio10}, while sudden changes in networks of computers are usually worth
    investigating for they might signal some external attack~\autocite{evolSecurity04}.
\end{itemize}

@@

Finally, we consider the case where there are more than two possible type of relations, which is
often called \emph{multilayer graphs}~\autocites{Kivela2014}{multiSurvey14}. In addition, we assume
that the graph is attributed, that is every node is described by a vector of numeric features called
the \emph{profile} of the node. Our assumption is then that links cannot be fully described by a
global homophily. Thus we take a more nuanced view, combining partial homophily (that is, nodes are
connected when they are similar on a subset of the attributes) and heterophily (that is, nodes are
connected when they are dissimilar on a subset of the attributes). As examples of the latter, think
of dating websites ---where most users are linked with users of the opposite
gender~\autocites{homophilyMyspace09}{Tinder16}; spreading innovation ---where meeting people with
different background and point of view is crucial to favor diversity and
creativity~\autocite{rogers2003diffusion}; and online news consumption ---where connecting people
from different sides of the political spectrum helps avoiding echo chambers and fueling a democratic
debate~\autocite{balancedNews17}. Our goal is to predict to which category a given belong. This is
widely applicable, for multilayer graphs are present  in various fields. In social networks, where
players of the Pardus massively multilayer online game can have six types of relations (friendship,
enmity, private message, trading, aggression and bounty)~\autocite{Szell2010} or where photo sharers
of the Flickr website can interact in eleven ways (either directly or through comments, tags, groups
and so on)~\autocite{RecoFlickrMulti11}. In citation networks, where authors of the DBLP dataset are
connected if they have co-authored a paper in one the \np{1000} conferences that form the edge
type~\autocite{communityDBLPbyConf05}, or where \np{5000} SIAM papers can be connected for five
reasons (abstract similarity, title similarity, keywords similarity, author similarity or
citation)~\autocite{articlesMultiSim11} in economic networks, where 951 ports are connected by 3
kind of ships (bulk dry carriers, container ships and oil tankers)~\autocite{ports3kindofships10} or
in the International Trade Network, where 162 countries are connected by 96 kind of commodities they
can exchange~\autocite{worldTradeNetwork10}. And in biology, where genes are connected in a
co-expression network under 130 different experimental conditions~\autocite{bioLayerExp11}.

There exists previous work that leans towards analyzing and characterising edges in such multilayer
graphs, although they do not share exact our goal of predicting category. For instance, one can
maximize the probability of a generative model in ego networks with $K$
categories~\autocites{LeskovecEgo12}{LeskovecEgo14}, look for communities that are both tightly
connected and share common keywords~\autocite{AttributedCommunity16}, find a partition maximizing
the density of intra community edges ---weighted by the endpoints profile similarities and a learned
community weight vector~\autocite{ZhangModelFree16}, recover with low distortion $K$ metric spaces
whose small-world superposition yield the observed graph~\autocite{Abraham2012a}, discover edge role
by computing high order topological features~\autocite{ahmed2017roles}, group nodes with common
attributes and extract frequent links between such group as edge type~\autocite{conceptualLinks12},
or cluster the profiles into a set of affine subspace of $\mathbb{R}^d$~\autocite{SCSurvey11}.

Our approach is to seek a small number of vectors that, once assigned to every edge, best explain
the graph (\ie maximize a score function between the edge vector and the profiles of its endpoints).
From this initial formulation, we derive two optimization problems, and initially solve them on
synthetic data, generated with a few natural topological constraints. We compared these solutions
with the result of a baseline approach that simply clusters the pairwise profile vectors using
$k$-means. As showed in \autoref{fig:edge}, on a graph with \np{500} nodes, \np{1300} edges and 7
types of edges (where each node is involved in at most 3 types of relation), we recover the hidden
edge assignment more accurately than the $k$-means baseline.

\begin{figure}[htpb]
  \centering
  \includegraphics[width=0.7\linewidth]{assets/raw/Gfixed_over0_step2_7dirs3_ambi.pdf}
  \caption{Distribution of the Adjusted Mutual Information (the closer to one the better) over
  \np{600} generation of 7 random directions, comparing our \textsf{optim} method with the $k$-means
baseline. \label{fig:edge}}
\end{figure}
